---
title: "excess_deaths2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r our world in data, include=FALSE}
library(dplyr)
library(readr)
#library(fpp2)
#library(forecast)
#library(tseries) # for adf.test() 
#library(FinTS) # for ArchTest()
#library(vrtest) # for Auto.AR()
#library(vars)
#library(MTS)
#library(tsbox)
library(plotly)
library(gghighlight)
library(tidyverse)
```

```{r load data, include=FALSE}
# load our world in data
owid_covid_data <- read.csv("data/owid-covid-data.csv")
covid_general <- owid_covid_data %>% 
  dplyr::select(iso_code, continent, location,date, total_vaccinations, people_vaccinated, stringency_index, population, population_density, median_age, aged_65_older, aged_70_older, gdp_per_capita, extreme_poverty, life_expectancy, cardiovasc_death_rate, diabetes_prevalence, handwashing_facilities, hosp_patients, hospital_beds_per_thousand)
# load climate data
climate_zones <- read.csv("data/climate zones - Munka1.csv") %>% dplyr::select(-1)
climate_zones[climate_zones$Country.Region=="US",]$Country.Region<-"United States"
names(climate_zones)[1] <- "location"
# World Value Survey
WVS <- read.csv2("data/WVSextra.csv")
names(WVS) <- c("location",'social_utility','conformity','trust','compliance')
# Legatum Institue's 2019 Prosperity Index
prosperity <-  read.csv("data/Prosperity Index_2019_Data_szerk.csv", sep=";")
names(prosperity)[1] <- "location"
url2 = "https://raw.githubusercontent.com/owid/covid-19-data/master/public/data/excess_mortality/excess_mortality.csv"
excess_mortality <- read_csv(url(url2))
```

```{r separate ts and non-ts, include=FALSE}
covid_general <- left_join(prosperity,covid_general, by=c("location"="location"))
covid_general <- left_join(covid_general, climate_zones, by=c("location"="location"))
length(unique(covid_general$location))
# filter our ts object
covid_general_ts <- covid_general %>% dplyr::select(location, iso_code, continent,date, stringency_index, total_vaccinations,people_vaccinated, hosp_patients)

#covid_general <- left_join(WVS, covid_general, by=c("location"="location"))
# number of countries
length(unique(covid_general$location)) #60
# filter out non ts objects
covid_no_ts <- covid_general %>% dplyr::select(-c("date","stringency_index","total_vaccinations","people_vaccinated","hosp_patients","handwashing_facilities","hospital_beds_per_thousand")) %>% distinct()
```

```{r World Mortality Dataset, eval=FALSE, echo=FALSE, include=FALSE}
# World Mortality Dataset
url = "https://raw.githubusercontent.com/akarlinsky/world_mortality/main/world_mortality.csv"
excess_death <- read_csv(url(url)) 
excess_death_sub <- excess_death %>%  filter(country_name %in% covid_general$location) #filter out selected country --> 90 left
length(unique(excess_death$country_name)) #116 countries
monthly <- excess_death %>% filter(time_unit=="monthly")
weekly <- excess_death %>% filter(time_unit=="weekly")
# lastest updated time by country
latest_date <- excess_death %>% 
  group_by(country_name) %>% 
  summarize(end_year=max(year),
            end_time=max(excess_death[excess_death$country_name==country_name & excess_death$year==end_year,]$time),
            after_covid_end_time = max(excess_death[excess_death$country_name==country_name & excess_death$year==2019,]$time),
            frequency=unique(excess_death[excess_death$country_name==country_name,]$time_unit))
```

```{r excess mortality, eval=FALSE, echo=FALSE, include=FALSE}
excess_mortality_timeseries <- read_csv("excess-mortality-main/excess-mortality-timeseries.csv")
em <- excess_mortality_timeseries %>%  filter(country_name %in% covid_general$location)
length(unique(em$country_name)) #91 countries

# lastest updated time by country
em_latest_date <- em %>% 
  group_by(country_name) %>% 
  summarize(end_year=max(year),
            end_time=max(em[em$country_name==country_name & em$year==end_year,]$time),
            frequency=unique(em[em$country_name==country_name,]$time_unit))
```

```{r function to get time series by country (weekly), eval=FALSE, echo=FALSE}
get_weekly_time_series <- function(country){
  country_subset <- weekly %>% filter(country_name == country)
  min_year = min(country_subset$year)
  min_time = min(country_subset[country_subset$year==min_year,]$time)
  all_ts <- ts(country_subset$deaths, frequency = 52,start = c(min_year,min_time)) # transform to time series (all time period)
  return(all_ts)
}
```

```{r test for US normality, eval=FALSE, echo=FALSE, include=FALSE}
us_ts = get_weekly_time_series("United States")
test_normality(us_ts,"US")
 # p < .05 --> not normal
```

```{r test for stationarity, eval=FALSE, echo=FALSE,include=FALSE}
us_ts2 <- BoxCox(us_ts,BoxCox.lambda(us_ts)) #not normal so need boxcox transformation
test_weekly_stationarity(us_ts2,"US")
# P <.05 --> stationary
# slow tapering in ACF
# p=2, P=0
```

```{r select ARIMA or SARIMA, eval=FALSE, echo=FALSE,include=FALSE}
us_sub <- window(us_ts,end=c(2019,52)) # to estimate the expected deaths, we do not yet consider time periods after COVID-19
(arima_us <- Arima(us_sub, 
                   order = c(2,0,0),
                   seasonal = list(order=c(0,1,0),
                                   period=52))) #AIC=3449.19
checkresiduals(arima_us)
(sarima_us <- auto.arima(us_sub, trace = F,
                          stepwise = T, #for faster stepwise selection
                          seasonal = T)) # allows for seasonal models
#SARIMA(1,0,0)(1,1,0)[52]: AIC 3406.46
checkresiduals(sarima_us) #Ljung p<.05
```

```{r, eval=FALSE, echo=FALSE,include=FALSE}
length(us_ts)-length(us_sub)
expected_us <- forecast(sarima_us,h=111)
expected_us_ts <- expected_us$mean
us_sub2 <- window(us_ts,start=c(2020,1))
excess_us <- us_sub2 - expected_us_ts
autoplot(expected_us, main="US Expected Mortality: ARIMA(2,0,2)(0,1,1)[52]", xlab="Year",series="Expected Deaths",ylab="Number of Deaths")+autolayer(us_sub2, series = "Reported Deaths")+autolayer(expected_us$mean, series="Expected Deaths")+theme_minimal()
```

## clustering excess deaths alone (using Euclidean distance)

```{r load excess mortality data from owid, include=FALSE}
excess_mortality <- excess_mortality %>% 
  mutate(month = as.numeric(substr(date,6,7)),
         year = as.numeric(substr(date,1,4)),
         year_month = substr(date,1,7))

# transfer all weekly to monthly (aggregate)
excess_mortality2 <- excess_mortality %>% 
  filter(location %in% covid_general$location) %>% 
  dplyr::select(location, date, year_month, year, month, excess_proj_all_ages,deaths_2020_2022_all_ages, projected_deaths_2020_2022_all_ages) %>%
  group_by(location, year_month, year, month) %>% 
  summarise(excess_deaths = sum(excess_proj_all_ages),
            reported_deaths = sum(deaths_2020_2022_all_ages),
            projected_deaths = sum(projected_deaths_2020_2022_all_ages),
            .groups="keep") %>% 
  drop_na() %>% 
  mutate(time_unit = as.numeric(sub("-",".",year_month)),
         p_score = excess_deaths/projected_deaths)
```

```{r filter out latest date, include=FALSE}
library(zoo) # for date operations
update_time = excess_mortality2 %>% group_by(location) %>% summarise(latest_date = max(time_unit),earliest = min(time_unit)) # choose update time since 2020/01 until 2021/12
selected_country = (update_time %>% filter(latest_date >= 2021.12 & earliest == 2020.01))$location

# keep all selected countries, filter out data after 2021/12
ex_country <- excess_mortality2 %>% 
  filter(location %in% selected_country & time_unit <= 2021.12) %>% 
  arrange(time_unit) %>% 
  mutate(date = as.Date(as.yearmon(year_month,"%Y-%m")))
```

### Plotting excess deaths in all country

-   69 countries were included after merging with other available datasets

-   Last updated date was selected to be December 2021

-   All excess deaths by country were aggregated into monthly values

```{r plot gghighlight, echo=FALSE}
ggplot(ex_country, aes(date, excess_deaths,color=location)) +
  geom_line(stat="identity") +
  ylab("Projected Excess Deaths") +
  gghighlight(max(excess_deaths) > 90000,
              max_highlight = 4,
              use_direct_label = TRUE) +
  theme_minimal() +
  theme(legend.position = 'none')
```

```{r plot brazil}
ggplot(ex_country[ex_country$location=="Brazil",], aes(date, excess_deaths)) +
  geom_line(stat="identity") +
  ylab("Projected Excess Deaths")+
  theme_minimal() 
```


```{r plot p-score}
ggplot(ex_country, aes(date, p_score,color=location)) +
  geom_line(stat="identity") +
  ylab("P-score of Excess Deaths") +
  gghighlight(max(p_score) > 2,
              max_highlight = 4,
              use_direct_label = TRUE) +
  theme_minimal() +
  theme(legend.position = 'none')
```


```{r long to wide, include=FALSE}
# exclude France because missing data from 2020/05 - 2021-10
ex_per_country <- ex_country[ex_country$location != "France",c("excess_deaths","time_unit","location")]
  
# long to wide
ex_per_country <- ex_per_country %>% 
  spread(location, excess_deaths)
```

### Fitting Hierarchical clustering using Euclidean distance

```{r fit hierarchical cusltering, include=FALSE}
# transpose excess deaths to matrix
deaths <- t(ex_per_country[-1])
deaths_dist <- proxy::dist(deaths, method="Euclidean")
ex_cluster_fit <- hclust(deaths_dist, method = "ward.D")
```

### Clustered Excess Deaths Dendrogram (Euclidean)

```{r plot dendrogram, include=FALSE}
# plot clusters
ggdendro::ggdendrogram(ex_cluster_fit, rotate=TRUE, theme_dendro = FALSE) + theme_minimal() + xlab("") + ylab("")
```

```{r merge cluster to full data, include=FALSE}
#assign the four clusters to the data using cutree() 
clustered_ex <- cutree(ex_cluster_fit, k=4)
clustered_exdata <- as.data.frame(as.table(clustered_ex))
colnames(clustered_exdata) <- c("location","cluster")
clustered_exdata$location <- as.character(clustered_exdata$location)
joined_clusters <- ex_country %>% 
  inner_join(clustered_exdata, by = "location")
```

```{r plot excess deaths cluster2, include=FALSE}
# cluster1
cluster1 <-
ggplot(joined_clusters %>% filter(cluster == "1") , aes(date, excess_deaths)) +
  geom_line(color="grey") +
  theme_minimal() +
  ylab("excess deaths") + xlab("") +
  geom_smooth(method="auto",color="red", se=F, size=0.5) +
  facet_wrap(~location, scales = "free")
# cluster2
cluster2 <- joined_clusters %>% filter(cluster == "2") %>% 
ggplot(aes(date, excess_deaths)) +
  geom_line(color="grey") +
  theme_minimal() +
  ylab("excess deaths") + xlab("") +
  geom_smooth(method="auto",color="red", se=F, size=0.5) +
  facet_wrap(~location)
# cluster3
cluster3 <- joined_clusters %>% filter(cluster == "3") %>% ggplot(aes(date, excess_deaths)) +
  geom_line(color="grey") +
  theme_minimal() +
  ylab("excess deaths") + xlab("") +
  geom_smooth(method="auto",color="red", se=F, size=0.5) +
  facet_wrap(~location) 
# cluster4
cluster4 <- joined_clusters %>% filter(cluster == "4") %>% ggplot(aes(date, excess_deaths)) +
  geom_line(color="grey") +
  theme_minimal() +
  ylab("excess deaths") + xlab("") +
  geom_smooth(method="auto",color="red", se=F, size=0.5) +
  facet_wrap(~location)
```

### Plot each cluster

```{r plot excess deaths cluster3, include=FALSE}
# how many countries per cluster
table(clustered_exdata$cluster)
cluster1
cluster2
cluster3
cluster4
```

## Clustering Excess Deaths using Dynamic Time Warp (DTW)

### Fit dtw hierarchical clustering
```{r long to wide2, include=FALSE}
# exclude France because missing data from 2020/05 - 2021-10
ex_per_country <- ex_country[ex_country$location != "France",c("p_score","time_unit","location")]
  
# long to wide
ex_per_country <- ex_per_country %>% 
  spread(location, p_score)
```

### Fitting Hierarchical clustering using Euclidean distance

```{r dtw hierarchical clustering}
deaths <- t(ex_per_country[-1])
# normalize data
deaths.norm <- BBmisc::normalize(deaths, method="standardize")
deaths_dist_norm <- dtw::dtwDist(deaths.norm) # calculate dtw distance
deaths_dist_norm <- as.dist(deaths_dist_norm) # convert to dist object
ex_cluster_fit2 <- hclust(deaths_dist_norm, method = "ward.D")

ex_dendrogram<- ggdendro::ggdendrogram(ex_cluster_fit2, rotate=TRUE, theme_dendro = FALSE) + theme_minimal() + xlab("") + ylab("")
ex_dendrogram
```

```{r pick number of cluster for excess deaths}
library(cluster)
deaths_sil_width <- c(NA)
for(i in 2:8){  
  deaths_pam_fit <- pam(deaths_dist_norm, diss = TRUE, k = i)  
  deaths_sil_width[i] <- deaths_pam_fit$silinfo$avg.width  
}
plot(1:8, deaths_sil_width,
     xlab = "Number of clusters",
     ylab = "Silhouette Width")
lines(1:8, deaths_sil_width)
```

```{r merge cluster to ex_country data, include=FALSE}
#assign the four clusters to the data using cutree() 
clustered_ex2 <- cutree(ex_cluster_fit2, k=6)

clustered_exdata2 <- as.data.frame(as.table(clustered_ex2))

colnames(clustered_exdata2) <- c("location","cluster")
# how many countries per cluster
table(clustered_exdata2$cluster)

clustered_exdata2$location <- as.character(clustered_exdata2$location)

joined_clusters2 <- ex_country %>% 
  inner_join(clustered_exdata2, by = "location")
```

```{r count by ex class}
joined_clusters2[,c("location","cluster")] %>% 
  distinct() %>% 
  mutate(cluser=as.factor(cluster)) %>% 
  group_by(cluster) %>% 
  summarise(count=n())
```


### plot each cluster

```{r dtw excess deaths cluster 1}
ex_cluster1 <- joined_clusters2 %>% filter(cluster == "1") %>% 
  ggplot(aes(date, p_score)) +
  geom_line(color="grey") +
  theme_minimal() +
  ylab("excess deaths") + xlab("") +
  geom_smooth(method="auto",color="red", se=F, size=0.5) +
  facet_wrap(~location, scales="free",ncol = 5)+
  ggtitle("P-scores of Excess Deaths Cluster 1")+
  theme(strip.text.y = element_text(size=2),
        strip.text.x = element_text(size=0.5))
ggplotly(ex_cluster1)
```

```{r dtw excess deaths cluster 2}
ex_cluster2 <- joined_clusters2 %>% filter(cluster == "2") %>% 
  ggplot(aes(date, p_score)) +
  geom_line(color="grey") +
  theme_minimal() +
  ylab("excess deaths") + xlab("") +
  geom_smooth(method="auto",color="red", se=F, size=0.5) +
  facet_wrap(~location, scales="free")+
  ggtitle("P-scores of Excess Deaths Cluster 2")
ggplotly(ex_cluster2)
```

```{r dtw excess deaths cluster 3}
ex_cluster3 <- joined_clusters2 %>% filter(cluster == "3") %>% 
  ggplot(aes(date, p_score)) +
  geom_line(color="grey") +
  theme_minimal() +
  ylab("excess deaths") + xlab("") +
  geom_smooth(method="auto",color="red", se=F, size=0.5) +
  facet_wrap(~location, scales="free")+
  ggtitle("P-scores of Excess Deaths Cluster 3")
ggplotly(ex_cluster3)
```

```{r dtw excess deaths cluster 4, eval=FALSE}
ex_cluster4 <- joined_clusters2 %>% filter(cluster == "4") %>% 
  ggplot(aes(date, excess_deaths)) +
  geom_line(color="grey") +
  theme_minimal() +
  ylab("excess deaths") + xlab("") +
  geom_smooth(method="auto",color="red", se=F, size=0.5) +
  facet_wrap(~location, scales = "free")+
  ggtitle("Excess Deaths Cluster 4")
ggplotly(ex_cluster4)
```

```{r dtw excess deaths cluster 5, eval=FALSE}
ex_cluster5 <- joined_clusters2 %>% filter(cluster == "5") %>% 
  ggplot(aes(date, excess_deaths)) +
  geom_line(color="grey") +
  theme_minimal() +
  ylab("excess deaths") + xlab("") +
  geom_smooth(method="auto",color="red", se=F, size=0.5) +
  facet_wrap(~location, scales = "free")+
  ggtitle("Excess Deaths Cluster 5")
ggplotly(ex_cluster5)
```

```{r dtw excess deaths cluster 6, eval=FALSE}
ex_cluster6 <- joined_clusters2 %>% filter(cluster == "6") %>% 
  ggplot(aes(date, excess_deaths)) +
  geom_line(color="grey") +
  theme_minimal() +
  ylab("excess deaths") + xlab("") +
  geom_smooth(method="auto",color="red", se=F, size=0.5) +
  facet_wrap(~location, scales = "free")+
  ggtitle("Excess Deaths Cluster 6")
ggplotly(ex_cluster6)
```

```{r merge excess deaths class to original data,include=FALSE}
ex_class <- joined_clusters2 %>% 
  group_by(location) %>% 
  summarise(excess_death_class = unique(cluster))

covid_no_ts_all <- left_join(ex_class, covid_no_ts,by = "location")
```

## clustering policy stringency alone

```{r data management for stringency data, include=FALSE}
stringency <- covid_general_ts %>% dplyr::select(location, date, stringency_index) %>% 
  filter(location %in% ex_class$location) %>% 
  mutate(year_month = substr(date,1,7),
         time_unit = as.numeric(sub("-",".",year_month))) %>% 
  drop_na()
# look at earliest and latest time in all country
summary((stringency %>% 
  filter(!is.na(stringency_index)) %>% 
  group_by(location) %>% 
  summarise(latest_date = max(date),
            earliest_date = min(date))))
### --> choose 2020-03-18 as earliest date and 2022-01-24 as latest
# compute stringency level by month (take the average)
month_stringency <- stringency %>% 
  group_by(location, year_month,time_unit) %>% 
  summarise(monthly_stringency = mean(stringency_index),
            .groups = "keep") %>% 
  # filter out date after 2021/12
  filter(time_unit <= 2021.12) %>% 
  arrange(time_unit) %>% 
  mutate(date = as.Date(as.yearmon(year_month,"%Y-%m")))
```

### Plotting policy stringency index by country

```{r stringency plot gghighlight, echo=FALSE}
ggplot(month_stringency, aes(date, monthly_stringency,color=location)) +
  geom_line(stat="identity") +
  ylab("Monthly Policy Stringency Index") +
  gghighlight(max(monthly_stringency) > 95,
              max_highlight = 4,
              use_direct_label = TRUE) +
  theme_minimal() +
  theme(legend.position = 'none')
```

### Fit dtw hierarchical clustering and plot dendrogram

```{r dtw hierarchical clustering for stringency}
# long to wide
stringency_per_country <- month_stringency[,c("monthly_stringency","time_unit","location")] %>% 
  spread(location, monthly_stringency) %>% 
  drop_na()
policy <- t(stringency_per_country[-1])
# normalize data
policy.norm <- BBmisc::normalize(policy, method="standardize")
policy_dist_norm <- dtw::dtwDist(policy.norm) # calculate dtw distance
policy_dist_norm <- as.dist(policy_dist_norm) # convert to dist object
policy_cluster_fit <- hclust(policy_dist_norm, method = "ward.D")
policy_dendrogram <- ggdendro::ggdendrogram(policy_cluster_fit, rotate=TRUE, theme_dendro = FALSE) + theme_minimal() + xlab("") + ylab("")
ggplotly(policy_dendrogram)
```

```{r pick number of cluster for policy}
policy_sil_width <- c(NA)
for(i in 2:8){  
  policy_pam_fit <- pam(policy_dist_norm, diss = TRUE, k = i)  
  policy_sil_width[i] <- policy_pam_fit$silinfo$avg.width  
}
plot(1:8, policy_sil_width,
     xlab = "Number of clusters",
     ylab = "Silhouette Width")
lines(1:8, policy_sil_width)
```


```{r merge stringency cluster to full data, include=FALSE}
#assign the four clusters to the data using cutree() 
clustered_policy <- as.data.frame(as.table(cutree(policy_cluster_fit, k=3)))
colnames(clustered_policy) <- c("location","cluster")
# how many countries per cluster
table(clustered_policy$cluster)
clustered_policy$location <- as.character(clustered_policy$location)
joined_clusters_policy <- month_stringency %>% 
  inner_join(clustered_policy, by = "location")
```

```{r count by policy class}
joined_clusters_policy[,c("location","cluster")] %>% 
  distinct() %>% 
  mutate(cluser=as.factor(cluster)) %>% 
  group_by(cluster) %>% 
  summarise(count=n())
```

### Plot policy stringency clusters

```{r policy cluster 1}
policy_cluster1 <-
  joined_clusters_policy %>% filter(cluster == "1") %>% 
  ggplot(aes(date, monthly_stringency)) +
  geom_line(color="grey") +
  theme_minimal() +
  ylab("policy stringency") + xlab("") +
  geom_smooth(method="auto",color="red", se=F, size=0.5) +
  facet_wrap(~location, scales="free")+
  ggtitle("Policy Stringency Cluster 1")
ggplotly(policy_cluster1)
```

```{r policy cluster 2}
policy_cluster2 <-
  joined_clusters_policy %>% filter(cluster == "2") %>% 
  ggplot(aes(date, monthly_stringency)) +
  geom_line(color="grey") +
  theme_minimal() +
  ylab("policy stringency") + xlab("") +
  geom_smooth(method="auto",color="red", se=F, size=0.5) +
  facet_wrap(~location, scales="free")+
  ggtitle("Policy Stringency Cluster 2")
ggplotly(policy_cluster2)
```

```{r policy cluster 3}
policy_cluster3 <-
  joined_clusters_policy %>% filter(cluster == "3") %>% 
  ggplot(aes(date, monthly_stringency)) +
  geom_line(color="grey") +
  theme_minimal() +
  ylab("policy stringency") + xlab("") +
  geom_smooth(method="auto",color="red", se=F, size=0.5) +
  facet_wrap(~location, scales="free")+
  ggtitle("Policy Stringency Cluster 3")
ggplotly(policy_cluster3)
```

```{r policy cluster 4, eval=FALSE}
policy_cluster4 <-
  joined_clusters_policy %>% filter(cluster == "4") %>% 
  ggplot(aes(date, monthly_stringency)) +
  geom_line(color="grey") +
  theme_minimal() +
  ylab("policy stringency") + xlab("") +
  geom_smooth(method="auto",color="red", se=F, size=0.5) +
  facet_wrap(~location, scales="free")+
  ggtitle("Policy Stringency Cluster 4")
ggplotly(policy_cluster4)
```


```{r merge stringency class to original data, eval=FALSE}
policy_class <- joined_clusters_policy %>% 
  group_by(location) %>% 
  summarise(stringency_class = unique(cluster))

covid_no_ts_all2 <- left_join(policy_class, covid_no_ts_all, by = "location")
write.csv(covid_no_ts_all2,"data/covid_no_ts_all.csv")
```

```{r merge p-score cumulative, eval=FALSE}
p_score_cum <- ex_country %>% 
  group_by(location) %>% 
  summarise(cum_p_score = sum(excess_deaths)/sum(reported_deaths))

continent <- covid_general %>% 
  select(location, iso_code, continent) %>% 
  distinct()

p_score_continent <- left_join(p_score_cum, continent, by='location')
write.csv(p_score_continent, "data/p_score_continent.csv")
write.csv(p_score_cum, "data/p_score_cum.csv")
```

